---
title: "Hepatocarcinoma Multi-Omics Differential Analysis of Sex, M v F"
author: "Robert M Flight"
date: last-modified
date-format: YYYY-MM-DD HH:mm
format: 
  html:
    embed-resources: true
    toc: true
  docx:
    keep-md: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
## target knits qmds in their own session, so load libraries here.
targets::tar_source("./packages.R")
tar_source("./R")
```

```{r}
#| label: load-targets
#| include: false
tar_load(c(
  bioamines_compare_sex,
  pm_compare_sex,
  lipidomics_compare_sex,
  rna_compare_sex,
  rna_qcqa,
  primary_metabolism_qcqa,
  lipidomics_qcqa,
  bioamines_qcqa
))
```

## Purpose

Differential analyses of transcriptomics and metabolomics (bioamines, primary metabolism, lipidomics) for several patients with hepatocarcinoma, comparing male vs female patients, and comparing with previous differential results for hepatocarcinoma.

There is an [Executive Summar](#executive-summary) at the end of this document.

## Data

Transcriptomics and metabolomics datasets from hepatocarcinoma.
Transcriptomics samples are normalized and statistical results reported by {DESeq2}.
Each metabolomics type were analyzed separately.
Metabolomics samples are normalized and statistical results reported by {DESeq2}, using a `local` fit for the modeling of dispersions of variance to mean abundance.

For all datasets, we merged the two replicate samples normal adjacent samples from Patient 6, by adding the abundances via {DESeq2}.

Significant in all of these is using a cutoff of adjusted-p-value $\le$ 0.01.

## Results

Sex doesn't seem to be a major factor, based on the PCA plots, and the number of significant transcripts or metabolites compared to cancer vs adjacent-normal.

## Transcriptomics

Number of features with an adjusted-p-value $\le$ 0.01:

* $\text{f}(\text{cancer} - \text{non-cancer}) - \text{m}(\text{cancer} - \text{non-cancer})$: 37;
* Cancer vs non-cancer: 1679;
* Common: 10.

```{r}
#| label: fig-transcriptomics-diff-sex-cancer
#| fig-cap: Upset plot comparing the number of significant differential transcripts for sex and cancer vs non-cancer.
rna_compare_sex
```



```{r}
#| label: fig-transcriptomics-pca-sex
#| fig-cap: PCA of transcriptomics data without outliers, with samples colored by sex.
rna_qcqa$pca_nooutlier_sex
```


## Bioamines

Number of features with an adjusted-p-value $\le$ 0.01:

* $\text{f}(\text{cancer} - \text{non-cancer}) - \text{m}(\text{cancer} - \text{non-cancer})$: 2;
* Cancer vs non-cancer: 1649;
* Common: 1.

```{r}
#| label: fig-bioamines-diff-sex-cancer
#| fig-cap: Upset plot comparing the number of significant differential bioamines for sex and cancer vs non-cancer
bioamines_compare_sex
```


```{r}
#| label: fig-bioamines-pca-sex
#| fig-cap: PCA of bioamines data without outliers, with samples colored by sex.
bioamines_qcqa$pca_nooutlier_sex
```



## Lipidomics

Number of features with an adjusted-p-value $\le$ 0.01:

* $\text{f}(\text{cancer} - \text{non-cancer}) - \text{m}(\text{cancer} - \text{non-cancer})$: 2;
* Cancer vs non-cancer: 292;
* Common: 2.

```{r}
#| label: fig-lipidomics-diff-sex-cancer
#| fig-cap: Upset plot comparing the number of significant differential lipidomics for sex and cancer vs non-cancer
lipidomics_compare_sex
```


```{r}
#| label: fig-lipidomics-pca-sex
#| fig-cap: PCA of lipidomics data without outliers, with samples colored by sex.
lipidomics_qcqa$pca_nooutlier_sex
```



## Primary Metabolites

Number of features with an adjusted-p-value $\le$ 0.01:

* $\text{f}(\text{cancer} - \text{non-cancer}) - \text{m}(\text{cancer} - \text{non-cancer})$: 1;
* Cancer vs non-cancer: 47;
* Common: 0.

```{r}
#| label: fig-pm-diff-sex-cancer
#| fig-cap: Upset plot comparing the number of significant differential primary metabolites for sex and cancer vs non-cancer
pm_compare_sex
```


```{r}
#| label: fig-pm-pca-sex
#| fig-cap: PCA of primary metabolites data without outliers, with samples colored by sex.
primary_metabolism_qcqa$pca_nooutlier_sex
```



## Executive Summary

* Sex is not a major contributor to the changes observed within patients between cancer and adjacent-normal tissue samples.
* Statistical results for female v male should have been provided in Excel files.